Untitled

### 2017-02-07 Manuscript/Manuscript_trends.Rmd

---
title: "Exploring how changes in age at death and cause of death have narrowed the black-white life expectancy gap over forty-five years in the United States."
output: html_document
---

##Riddell CA, Morrison KT, Kaufman JS, Harper S

###Abstract

###Introduction
The life expectancy of a population serves as a barometer for overall health. In 2014, life expectancy in the United States was 78.7 years, but hidden within this figure, are differences in life expectancy by race. Non-hispanic whites lived 3.6 years longer than non-hispanic blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white americans have experienced increases in life expectancy since X, the difference between their life expectancies has varied. During the 1980s, the life expectancy gap widened and then narrowed in the 1990s due to relative mortality improvements among blacks in deaths due to homicide, HIV, unintentional injuries, and, among females, heart disease [Harper 2007]. The gap continued to narrow during the 2000s, but states differed in the magnitude of improvement that was made. For example, the life expectancy gap in New York decreased from 8.0 years to 2.4 years between 1990 and 2009, but in California the decrease was much more modest -- from 6.7 years to 5.6 years [Harper 2014]. This study seeks to better understand these state-level differences, by examining how changes in age at death and cause of death over time have worked to narrow or exacerbate the life expectancy gap.

- fundamentally interested in the qu is there a common story across most states or are there important differences in what happened in NY vs CA vs TX - can we say anything about what we know about state-specific policies, for example CA has strong anti-smoking legislation. Does this help us understand why the gap in Wisconsin was increasing?

- is there a lot of heterogeneity or is there not? - same something about the proportionate contribution by COD and by age - e.g., changes in injuries accounting for X% for closing the gap in NY but in AL it was Y%.

###Methods

**The data**

We used software maintained by the National Cancer Institute's Surveillance, Epidemiology, and End Results Program (SEER*Stat) to extract data on the number of deaths and the population size in all 50 US states and the District of Columbia. The mortality data was originally collected by the National Center for Health Statistics (NCHS) from death certificates and stored by the National Vital Statistics System. The population data was derived by the NCHS using intercensal bridged-race population estimates. This study using data on the number of deaths and the population size according to race (black or white), sex (male or female), age group (<1 year old, 1-4 years old, followed by 5-year age groupings, with a final grouping of 85 or more years of age), year (1969 -- 2013), and cause of death. Cause of death was originally recorded using the International Classification of Diseases (ICD) codes versions 8 (1969-1979), 9 (1979-1998), and 10 (1999-2013). These codes were then grouped into 9X causes of death within the software, which we further aggregated into six broad causes of death for this analysis: Cancers, Cardiovascular disease, Communicable disease, Injuries, Non-communicable disease, and All other causes. This recoding is illustrated in eTable 1 in the eAppendix of the Supplement. When the number of deaths for a population is between 1 and 9, this data is suppressed by the software and produces a missing death count for that strata. However, death counts of exactly 0 or greater than 9 are known precisely.

- CR to add line about which states were not considered.

- CR to add line about hispanic origin.

**Statistical Methods**

The data produced two statistical issues that were addressed by our statistical methods. First, our methodology needed to impute the missing death counts. Second, several states had small blacks populations which produced even smaller population sizes in each sex-age category. In these cases, the estimation of the sex-age mortality rates is imprecise, and produces abrupt changes in the estimation of mortality rates over time that are unrealistic. To better estimate the underlying age-specific mortality rates, a method that produces smoother age-specific mortality trends was required.

To address these challenges, we used a Bayesian autoregressive model to smooth the data over time, and incorporated a truncated Poisson model to impute the suppressed death counts. Specifically, we used Poisson regression to model the number of deaths in each state-age-sex-race-year category as a function of the at-risk population. The mortality rates over time within each state-age-sex-race category are modelled autoregressively, where the log mortality rate in the current year was modelled as a normal variable centred at the log mortality rate estimated from the previous year. Truncated Poisson regression was used to impute the supressed death counts, by specifying that the supressed counts were sampled from a Poisson distribution and were constrained to be between 1 and the maximum of 9 and the at-risk population size (which may be smaller than 9). These models were run separately for every combination of state, race, and sex, and produced death counts that had been smoothed over time for every cause of death within each age group.

To calculate life expectancy at birth, the smoothed death counts were aggregated to estimate the total deaths in each age group and standard life table methods were applied [ref]. Life expectancy at birth (in years) was calculated for blacks and whites in each state and year according to sex. The difference in life expectancies between blacks and whites was calculated by subtracting the black life expectancy from the white life expectancy in every state, year, and sex.

To decompose the year- and state- specific difference in life expectancy between blacks and whites, we performed a decomposition analysis using the Arriaga method [ref]. First, the decomposition analysis partitions the difference in life expectancy by age group, which can be interpreted as the amount of the gap (in years) that can be attributed to higher age-specific mortality rates in blacks. For example, a decomposition of a gap of five years may find that two of the five years (40% of the gap) is attributable to a higher mortality rate in blacks in the youngest age group (< 1 years old) compared with the youngest whites. This age-specific decomposition can be further partitioned according to cause of death, producing the number of years of the gap that is attributable to every age- and cause-of-death combination. For example, a decomposition of a gap of 8 years may find that 1.3 years of this gap is attributable to more injury-related deaths in blacks aged 25-29 years old. We also summed these age- and cause-of-death partitions over all ages to calculate the marginal contribution of each cause-of-death to the difference.

In order to understand whether states fared similarly over time, we divided the time frame into three eras which roughly encapsulate an initial period of increases in life expectancy for both races and sexes (1969-1983), an intermediate period of decreases and stagnating life expectancy for blacks but not for whites (1983-1993) and a final period of increasing life expectancy for both races (1993-2013). While these eras represent different lengths of time, they best capture three distinct periods of changes to the difference in life expectancy: an intial narrowing of the gap, followed by a widening of the gap that diminished previous progress, and the most recent twenty years of sustained narrowing (eAppendix Figures 1 and 2).

**Will need to change the following**
Finally, we calculated the change in the life expectancy gap over time within each state-, race-, and sex-specific stratum. The change in the life expectancy gap is calculated by substracting the gap that occurred in the later year from the gap that occurred in the earlier year. This change in gap can be decomposed by age by taking the difference in the age-specfic contributions across the two years, and can similarly be calculated by cause.

All data analyses were performed using R version 3.3.1. The bayesian smoothing was performed using `JAGS` (version X.X), and the life expectnacy and decomposition analysis were performed using user-written functions found in the Supplement.

**Add in the stuff about credible intervals**

- sampled from posterior distribution with thinning

###Results
```{r initialize, echo=FALSE, warning=F, message=F}
library(shiny)
library(plotly)
library(viridis)
library(scales)
library(shinythemes)
library(dplyr)
library(purrr)
library(stringr)
source("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Code/life_expectancy_functions.R")

age_cod_results <-rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"),
                        read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
                        )

age_cod_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_1.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_2.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_3.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_4.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_5.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_6.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_7.csv"),
                         read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_cod_set_8.csv")
                         )

age_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_3.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_4.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_5.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_6.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_7.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/age_decomp_set_8.csv")
                            )

BlackWhite_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_3.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_4.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_5.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_6.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_7.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/BlackWhite_set_8.csv")
                            )

cod_decomp_results <- rbind(read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_1.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_2.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_3.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_4.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_5.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_6.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_7.csv"),
                            read.csv("/Users/corinneriddell/Documents/repos/BlackWhiteMortalityGap/Results/cod_marginal_set_8.csv")
                            )
```

```{r tidy_data, echo=FALSE}

ncols <- str_split_fixed(as.character(BlackWhite_results$stratum.id), "\\.", 3)
BlackWhite_results$state <- as.factor(ncols[ , 1])
BlackWhite_results$year <- as.numeric(ncols[ , 2])
BlackWhite_results$sex <- as.factor(ncols[ , 3])

age_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_decomp_results, by = "stratum.id")
cod_decomp_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], cod_decomp_results, by = "stratum.id")
age_cod_results <- merge(BlackWhite_results[, c("stratum.id", "state", "year", "sex")], age_cod_results, by = "stratum.id")

COD.levels <- c("Cardiovascular", "Cancers", "Communicable", 
                "Non-communicable", "Injuries", "All other causes")
cod_decomp_results$COD <- factor(cod_decomp_results$COD, COD.levels)
age_cod_results$COD <- factor(age_cod_results$COD, COD.levels)
```

```{r add_Census_Region, echo=FALSE}
#note that the Census Region of the South is partitioned into two for better display.

BlackWhite_results$Census_Region <- "South"
BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
                                                 "Nebraska", "North Dakota", "South Dakota", 
                                                 "Illinois", "Indiana", "Michigan", "Ohio", 
                                                 "Wisconsin")] <- "Midwest"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
                                                 "New Hampshire", "Rhode Island", "Vermont",
                                                 "New Jersey", "New York", "Pennsylvania")] <- "Northeast"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
                                                 "Nevada", "New Mexico", "Utah", "Wyoming",
                                                 "Alaska", "California", "Hawaii", "Oregon",
                                                 "Washington")] <- "West"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                                                  "North Carolina", "South Carolina", 
                                                  "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"

BlackWhite_results$Census_Region[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee", 
                                                  "Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "South Central"

BlackWhite_results$Census_Region <- factor(BlackWhite_results$Census_Region, levels = c("West", "Midwest", "South Central", "South Atlantic", "Northeast"))
```

```{r add_Census_Division, ncol = 3, echo = F}
BlackWhite_results$Census_Division <- NA

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Connecticut", "Maine", "Massachusetts", 
                                                                   "New Hampshire", "Rhode Island", "Vermont")] <- "New England"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("New Jersey", "New York", "Pennsylvania")] <- "Mid-Atlantic"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Illinois", "Indiana", "Michigan", "Ohio", 
                                                                 "Wisconsin")] <- "East North Central"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Iowa", "Kansas", "Minnesota", "Missouri", 
                                                                 "Nebraska", "North Dakota", "South Dakota")] <- "West North Central"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Delaware", "Florida", "Georgia", "Maryland", 
                                                                 "North Carolina", "South Carolina", 
                                                                 "Virginia", "Washington DC", "West Virginia")] <- "South Atlantic"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alabama", "Kentucky", "Mississippi", "Tennessee")] <- "East South Central"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arkansas", "Louisiana", "Oklahoma", "Texas")] <- "West South Central"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Arizona", "Colorado", "Idaho", "Montana", 
                                                                 "Nevada", "New Mexico", "Utah", "Wyoming")] <- "Mountain"

BlackWhite_results$Census_Division[BlackWhite_results$state %in% c("Alaska", "California", "Hawaii", "Oregon","Washington")] <- "Pacific"

BlackWhite_results$Census_Division <- factor(BlackWhite_results$Census_Division, levels = c("New England", "Mid-Atlantic", "East North Central", "West North Central", "South Atlantic", "East South Central", "West South Central", "Mountain", "Pacific"))

```{r addvar_State_abbreviation, echo=FALSE}
BlackWhite_results <- BlackWhite_results %>% mutate(stabbrs = factor(state,  
                                                                     labels = c("AL", "AZ", "AR", "CA",
                                                                                "CO", "CT", "DE", "FL",
                                                                                "GA", "IL", "IN", "IA",
                                                                                "KS", "KY", "LA", "MD",
                                                                                "MA", "MI", "MN", "MS",
                                                                                "MO", "NE", "NV", "NJ", 
                                                                                "NM", "NY", "NC", "OH", 
                                                                                "OK", "OR", "PA", "RI", 
                                                                                "SC", "TN", "TX", "VA", 
                                                                                "WA", "DC", "WV", "WI")))

                                      # labels = c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "DC",
                                      #            "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", 
                                      #            "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", 
                                      #            "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", 
                                      #            "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", 
                                      #            "VT", "VA", "WA", "WV", "WI", "WY")))

cod_decomp_results <- merge(BlackWhite_results %>% select(stratum.id, Census_Region, Census_Division, stabbrs), 
                            cod_decomp_results, by = "stratum.id", all.y = T)
```

```{r prep_era_figure_males, echo = F}
year1 <- 1969
year2 <- 1983
year3 <- 1993
year4 <- 2013

bw_dat <- BlackWhite_results %>% 
  filter(year %in% c(year1, year2, year3, year4), is.na(LE_wbgap) == F) %>%
  select(state, year, LE_wbgap, sex) %>%
  arrange(year)

bw_dat <- tidyr::spread(bw_dat, year, LE_wbgap) 
names(bw_dat)[3:6] <- c("first.gap","second.gap", "third.gap", "fourth.gap")  
bw_dat.males <- bw_dat %>% filter(sex == "Male") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

bw_dat.females <- bw_dat %>% filter(sex == "Female") %>%
  mutate(era1.diff = first.gap - second.gap,
         era2.diff = second.gap - third.gap,
         era3.diff = third.gap - fourth.gap,
         state.gdiff.order.1 = reorder(state, first.gap),
         state.gdiff.num.1 = as.numeric(state.gdiff.order.1),
         state.gdiff.order.2 = reorder(state, second.gap),
         state.gdiff.num.2 = as.numeric(state.gdiff.order.2),
         state.gdiff.order.3 = reorder(state, third.gap),
         state.gdiff.num.3 = as.numeric(state.gdiff.order.3)
  )

cod.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     COD = factor(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

age.df <- data.frame(state = factor(), sex = factor(), era = character(), 
                     age_minbin = numeric(), Contribution.to.change = numeric(), 
                     Contrib.to.change.prop = numeric(), narrowed_gap = logical())

years <- c(1969, 1983, 1993, 2013)

for (state.i in levels(BlackWhite_results$state)) {
  for (sex.i in c("Male", "Female")) {
    for (year.i in 1:3) {

      table1 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(COD)
      table2 <- cod_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(COD)

      table3 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i]) %>% arrange(age_minbin)
      table4 <- age_decomp_results %>% filter(state == state.i & sex == sex.i & year == years[year.i + 1]) %>% arrange(age_minbin)

      temp1 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "COD", 
                                                     decomp.table1 = table1, 
                                                     decomp.table2 = table2, 
                                                     type.of.data = "results"))

      temp2 <- data.frame(state = state.i, sex = sex.i, era = paste0(years[year.i], "-",  years[year.i + 1]),
                          contribution.to.gap.change(type.of.decomp = "Age", 
                                                     decomp.table1 = table3, 
                                                     decomp.table2 = table4,
                                                     type.of.data = "results"))

      cod.df <- rbind(cod.df, temp1)
      age.df <- rbind(age.df, temp2)
    }
  }
}

```{r more_figure_prep, echo = F}

cod.df <- make_pretty_decomp_plot(decomp.table = cod.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "COD", type.of.data = "results")

age.df <- make_pretty_decomp_plot(decomp.table = age.df, strat.var.1 = "state", 
                        strat.var.2 = "sex", strat.var.3 = "era", 
                        sign.var = "narrowed_gap",
                        decomp.var = "Contribution.to.change",
                        decomp.var.prop = "Contrib.to.change.prop", 
                        partition.bar.var = "Ages", type.of.data = "results")

cod.df <- merge(cod.df, BlackWhite_results[BlackWhite_results$year == 1969 & BlackWhite_results$sex == "Male", 
                                           c("state", "Census_Region", "Census_Division", "stabbrs")], by = c("state"))
```

*Changes to the difference in life expectancy in US males*

**Figure 1a** displays each state's male life expectancy gap in 1969 (vertical dashed line) vs. their gap in 1983 (solid dashed line), with the change in the gap being represented by the number of years between these lines. The decomposition (in years) of the change in the gap by cause of death is also shown for each state.

```{r, echo = F}
sum1 <- BlackWhite_results %>% filter(year == 1969 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

diff <- bw_dat.males %>% arrange(era1.diff)

max1 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983"])
mc1 <- cod.df$COD[cod.df$state == diff[dim(diff)[1], ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max1]

max2 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983"])
mc2 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-1, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max2]

max3 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983"])
mc3 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-2, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max3]

max4 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983"])
mc4 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-3, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max4]

max5 <- max(cod.df$Contribution.to.change[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983"])
mc5 <- cod.df$COD[cod.df$state == diff[dim(diff)[1]-4, ]$state & cod.df$era == "1969-1983" & cod.df$Contribution.to.change == max5]
```

What we see so far (based on these `r dim(sum1)[1]` states):

- The gap in 1969 ranges from `r round(sum1[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum1[1, ]$LE_wbgap_lcl, 1)`, `r round(sum1[1, ]$LE_wbgap_ucl, 1)`] in `r sum1[1, ]$state ` to `r round(sum1[dim(sum1)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum1[dim(sum1)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum1[dim(sum1)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum1[dim(sum1)[1], ]$state `.

- For most states, the gap decreases between 1969 and 1983, with the largest decrease found in `r diff[dim(diff)[1], ]$state` (change in gap of `r round(diff[dim(diff)[1], ]$era1.diff, 1)`, `r round(max1,1)` of which is due to `r mc1`, 
`r diff[dim(diff)[1]-1, ]$state` (change of `r round(diff[dim(diff)[1]-1, ]$era1.diff, 1)`, `r round(max2,2)` due to `r mc2`), `r diff[dim(diff)[1]-2, ]$state` (change of `r round(diff[dim(diff)[1]-2, ]$era1.diff, 1)`, `r round(max3,1)` due to `r mc3`), `r diff[dim(diff)[1]-3, ]$state` (change of `r round(diff[dim(diff)[1]-3, ]$era1.diff, 1)`, `r round(max4,1)` due to `r mc4`), and `r diff[dim(diff)[1]-4, ]$state` (change of `r round(diff[dim(diff)[1]-4, ]$era1.diff, 1)`, `r round(max5,1)` due to `r mc5`).

- The gap increases for some states. Most notably in DC (CVD and other causes), Minnesota (CVD), Wisconsin (CVD), California (CVD). Also, for these states injuries do not contribute as much to narrowing the gap (and in Minnesota, injuries widens the gap), so their gap increasing appears to be due to a combination of changes to CVD making blacks relatively worse off and less changes to injuries than experienced by other states.

- Injuries across the board contribute to narrowing the gap over time, except for Minnesota and Nebraska.

- Communicable disease narrows (except Arizona and Minnesota), but much smaller contribution than injuries.

- Generally, all other causes narrows, again much smaller contribution than injuries. In some states its contribution is larger to the narrowing -- namely, New York, Missouria, Pennsylvania, Rhode Island, and Mississippi.

- Generally, CVD increases gap.

- Cancer increases the gap everywhere except New Mexico. Size of contribution to change is often smaller than contribution of CVD, but not always the case.

- Generally, non-communicable disease increases the gap but the contribution is tiny. Notable differences in Massachusetts, New York, Oregon, and Maryland where it decreases gap.

**Figure 1a: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1969 and 1983 by US state**

```{r figure1, fig.align = "center", echo=F}

cod.df.males <- merge(subset(cod.df, sex == "Male"), bw_dat.males, by = c("state"))
cod.df.males <- cod.df.males %>% mutate(new.start.1 = -start + first.gap, new.finish.1 = -finish + first.gap,
                                        new.start.2 = -start + second.gap, new.finish.2 = -finish + second.gap,
                                        new.start.3 = -start + third.gap, new.finish.3 = -finish + third.gap)

y.labels <- levels(factor(cod.df.males$state.gdiff.order.1))
y.num.breaks <- length(levels(factor(cod.df.males$state.gdiff.order.1)))

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
#    facet_grid(.~ Census_Region) +
    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
   # geom_segment(data = bw_dat.males,
  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
  #  geom_segment(data = bw_dat.males,
  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era1
```

```{r, echo =F}
sum2 <- BlackWhite_results %>% filter(year == 1983 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

**Figure 1b** displays each state's male life expectancy gap in 1983 (vertical dashed line) vs. their gap in 1993 (solid dashed line).

- Be careful with the interpretation: the x-axis is much wider for this plot because of DC. (Zoom into same x range to aid interpretation at first)

- The gap in 1983 ranges from `r round(sum2[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum2[1, ]$LE_wbgap_lcl, 1)`, `r round(sum2[1, ]$LE_wbgap_ucl, 1)`] in `r sum2[1, ]$state ` to `r round(sum2[dim(sum2)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum2[dim(sum2)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum2[dim(sum2)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum2[dim(sum2)[1], ]$state `.

- Except for New Mexico, Arizona, and Kentucky, West Virginia, and Oregon, the life expectancy gap increases in each state, with the biggest driver being communicable disease (HIV/AIDS), especially in MD, NY, NJ, and CT. Injuries also contribute to the gap widening over this time frame, most notably in Wisconsin, MD, and LA. Thus, compared to previous period, these two causes of death led to a widening rather than a narrowing of the gap.

- Similar to the firts time frame, CVD and cancer also contribute to widening the gap over time. The magnitude of the contribution of CVD is quite variable by state, with Mississippi, Minnesota, and Pennsylvania seeing the largest absolute contributions.

- Also similar is the minimal effect of  non-communicable disease on the gap, while all other causes contributes to narrowing the gap, with greatest contribution to narrowing in NY, Pennsylvania, CT, Mississippi, and FL.

- DC was especially hard hit, with injuries driving the increase in the gap from 10.5 years to 15.9 years in this 10-year period, an increase of X, Y of which can be attributed to injuries.

**Figure 1b: Decomposition of the change in the male black-white life expectancy gap by cause of death between 1983 and 1993 by US state**

```{r figure1b, echo = F}
y.labels.2 <- levels(factor(cod.df.males$state.gdiff.order.2))
y.num.breaks.2 <- length(levels(factor(cod.df.males$state.gdiff.order.2)))

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

males.era2 
```

```{r, echo =F}
sum3 <- BlackWhite_results %>% filter(year == 1993 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)

sum4 <- BlackWhite_results %>% filter(year == 2013 & sex == "Male") %>% 
  arrange(LE_wbgap) %>% 
  select(state, LE_wbgap_lcl, LE_wbgap, LE_wbgap_ucl)
```

- The gap in 1993 ranges from `r round(sum3[1, ]$LE_wbgap, 1)` [95% CI: `r round(sum3[1, ]$LE_wbgap_lcl, 1)`, `r round(sum3[1, ]$LE_wbgap_ucl, 1)`] in `r sum3[1, ]$state ` to `r round(sum3[dim(sum3)[1], ]$LE_wbgap, 1)` [95% CI: `r round(sum3[dim(sum3)[1], ]$LE_wbgap_lcl, 1)`, `r round(sum3[dim(sum3)[1], ]$LE_wbgap_ucl, 1)`] years in `r sum3[dim(sum3)[1], ]$state ` (or `r round(sum3[dim(sum3)[1]-1, ]$LE_wbgap, 1)` when DC is excluded).

>>>>

---
title: "Untitled"
author: "Corinne Riddell"
date: "February 7, 2017"
output: html_document
---
